---
title: "MENA Exceptionalism and BART"
author: "Daniel K Baissa"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this example I will use BART to predict coups (not the DV) globally and use it to explain my logic for looking at MENA exceptionalism

I will start by loading the libraries and setting up bartMachine

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(rJava)

options(java.parameters = "-Xmx50g")     
library(bartMachine)
set_bart_machine_num_cores(30)
```


This is just a subset of data for illustrative perpouses and speed on my laptop.

I will set up the data for BARTmachine

https://statisticsglobe.com/convert-factor-to-dummy-variables-in-r
```{r message=FALSE, warning=FALSE}
df <- read_csv("Data/Vdem_Banks.csv") |> 
  select(!c(suvival_time, coup_number, id, ...1)) |> 
  mutate(Region = as.factor(Region)) |> 
  filter(country_name != "Israel") # Removing Israel because its an outlier in the MENA





df <- data.frame(df[ , ! colnames(df) %in% "Region"],       # Create dummy data
                         model.matrix( ~ (Region) - 1, df) )


X <- df |> 
  na.omit()

y <- as.factor(X$e_pt_coup)

X$e_pt_coup <- NULL

y <- relevel(y, "1")

X <- as.data.frame(X)
```

Here are the variables on the right hand side of this model:

```{r paged.print=TRUE}
# Convert column names to a table for ease of reading
col_table <- matrix(colnames(X), ncol = 2, byrow = TRUE)

# Print the table
col_table
```


Now let's run a basic BART model. The data are extremely sparce so I will set prob_rule_class = .1 to compensate. This can be validated by looking at the confusion matrix to make sure the model has a good fit.

```{r message=FALSE, warning=FALSE}
bm <- bartMachine(y = y, X = X, prob_rule_class = .1)
```

```{r}
bm
```

Here we see that the model accurately classifies `r (1 - bm[["confusion_matrix"]][["model errors"]][2]) *100`\% of non-coups and `r (1-bm[["confusion_matrix"]][["model errors"]][1]) *100`\% of coups with an overall accuracy rate of `r (1-bm[["confusion_matrix"]][["model errors"]][3]) *100`\%


## Variable Selection

Note that there are `r length(bm$training_data_features)` variables in the final model. Let's select a smaller number of variables to allow for an easier comparison to see how Region might differ.

```{r}
vs <-  var_selection_by_permute(bm, plot  = FALSE)

```

Now we will use local names to select the following `r vs$important_vars_local_names` as variables


```{r}
X2 <- df |> 
  select(c(Region3, e_pt_coup, vs$important_vars_local_names)) |> 
  na.omit()

y2 <- X2$e_pt_coup

X2$e_pt_coup <- NULL


bm2 <- bartMachine(y = y2, X = X2, prob_rule_class = .1)

```


## PD Plot

Region 3 is for the MENA

```{r}

pd_plot(bm2,  "Region3")
```

Here we find that the MENA region is about the same with respect to coup chance as any other region.

```{r}
for (i in 1:length(bm2$training_data_features)) {
  pd_plot(bm2,  bm2$training_data_features[i])

}


```

Here we can see that Party Competition has a U shaped Function. Democracy is negativly associated with coups, Infant mortality is poitivly assoicated with coups, and political polarization is moderatly associated with coups at the extremel levels. 

## Within the MENA

Now let's look at the results within the MENA... This unfortunately would be like what we did before, because we want to know about variation within the region. Hariri used a 2 stage model to say that MENA is unique. I am not aware of such a BART model today. But I am not sure what he did made sense anyways.


### Countries in the MENA dataset

```{r}
df |> 
  filter(Region3 == 1) |> 
  select(country_name) |> 
  unique()


```

### Stage 1

Stage 1 would be something like starting by predicting the MENA region. Technically that would be finding everything that is related to being in the MENA region. 



```{r}
X3 <- df |> 
  select(-c("Region1", 
            "Region2",
            "Region4",
            "Region5",
            "Region6",
            "Region7",
            "Region8",
            "Region9",
            "Region10",
            "country_name")) |> 
  na.omit()

y3 <- as.factor(X3$Region3)
y3 <- relevel(y3, "1")

X3$Region3 <- NULL

bm3 <- bartMachine(y = y3, X = X3, prob_rule_class = .5)

bm3
```

This model was extremely effetive at predicting MENA countries.


### Predictors of the MENA

Now what variables predict being in the MENA?

```{r}
vs <-  var_selection_by_permute(bm3, plot  = FALSE)

```

The following variables predict being a MENA country`r vs$important_vars_local_names`


### Effect of Predictors of MENA

Now let's see how those predictors determine impact the MENA region


```{r}

X4 <- df |> 
  select(c(Region3, vs$important_vars_local_names)) |> 
  na.omit()

y4 <- as.factor(X4$Region3)
y4 <- relevel(y4, "1")

X4$Region3 <- NULL

bm4 <- bartMachine(y = y4, X = X4, prob_rule_class = .5)

bm4





for (i in 1:length(bm4$training_data_features)) {
  pd_plot(bm4,  bm4$training_data_features[i])

}
```


This findings indicate that the MENA countries have lower party competition across their regions in their countries, Higher GDP Per Capitas, though not the highest. Produce more oil per capita, and are more likely to have their HOG propose legislation in practice (v2exdfpphg) though that may not be signficant. Oddly, it looks like V-dem also codes the MENA countries as having fewer barriers to political parties? 

